Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-31 Report/Sections/results.tex
Stylistic changes to report

After fitting the four candidate models to simulated data, we fit the same four models to the data shown in Figure \ref{fig:data}. Here we report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and asses the quality of the fit. We also compare the results of the CarHHMM-DFT with the other candidate models.

The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to these short dives such as resting, traveling, and searching for prey \citep{Tennessen:2019a}.
Dive type 2 is longer and deeper. These types of deep sustained dives have been associated with behaviours such as foraging and socialization in killer whales \citep{Tennessen:2019a,Tennessen:2019b}. However, \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m, so it is unlikely that any of the 267 dives used in this study are successful foraging dives.

The three subdive behaviours have distinct $Z^{*(2)}$ values, as the means of each subdive state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean value of $Z^{*(2)}$ and the smallest variance corresponding to $Z^{*(1)}$. It also has the highest auto-correlation in $Z^{*(1)}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower in this state. We hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $Z^{*(2)}$ and variance of $Z^{*(1)}$ in subdive state 3 is much higher than the other two states. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between consecutive segments. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions from dive to dive. For dive types 1 and 2,
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for the fine-scale within-dive transitions.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that most dives are short type 1 dives. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short dives before a killer whale performs a longer, deeper dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives. Using less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of 6 selected dives. The supplementary material also shows the estimated probability of each dive and subdive state given the data (Section S-2.2).

@@

Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms. A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}. To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution. Mathematically, the pseudo-residual of an observation $y_t$ in a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals of this model mostly support that the model is well-specified. One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution.

We also plot histograms of $Y$ and $Z^*$ separated by dive type or subdive state in Figure \ref{fig:empirical_dist}. Observations are weighted by the decoded probability that they correspond to a particular hidden state, and the histogram is finally plotted over the emission distribution learned by the CarHHMM-DFT (Figure \ref{fig:empirical_dist}). Our results again mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed. Another exception is $Z^{*(1)}$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}. See Section S-2.? for histograms for every observation and hidden state.

@@

The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale. However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives, and it therefore does not provide any information relating to the dive-level Markov chain or the relationship between it and the subdive level. For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.

The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3. In addition, the HHMM-DFT produces estimates of $\sigma_1^{*(1)}$, $\sigma_1^{*(2)}$, and $\sigma_1^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those for the CarHHMM-DFT across all axes of acceleration. The estimated uncertainties of the estimates of $\mu_1^{*(1)}$, $\mu_1^{*(2)}$, and $\mu_1^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration. These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation. The empirical auto-correlation of $Z^{*(1)}$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT show several outliers. Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.

The CarHHMM does not use the Fourier sums of the acceleration data, so it fails produce biologically interpretable results when decoding the subdive states. The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution. See section S-2 for a more complete set of results for each of the candidate models.

>>>>

We fit all four models from the simulation study to the data shown in Figure \ref{fig:data} to test their performance in a real-world setting. We first report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and then asses the quality of the fit. Finally, we compare the results of the CarHHMM-DFT with those of the other candidate models.

The emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, traveling, and searching for prey.
On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}. No dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m. It is therefore unlikely that any of the dives used in this study are successful foraging dives.

The three subdive behaviours have distinct average ``wiggliness" values, as the means of $Z^{*(2)}$ for each state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean of $Z^{*(2)}$ and the smallest variance corresponding to $Z^{*(1)}$. It also has the highest auto-correlation in $Z^{*(1)}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($Z^{*(2)}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower in than subdive state 1. We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $Z^{*(2)}$ and variance of $Z^{*(1)}$ in subdive state 3 is much higher than the other two states. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between segments. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions between dives. Within dives, we have
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for dive types 1 and 2.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that most dives are short, type 1 dives. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives. Using less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. The supplementary material also shows the estimated probability of each dive and subdive state given the data (Section S-2.2).

@@

Two visual tools are used to evaluate this model: pseudo-residuals and empirical histograms. A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}. To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution. Mathematically, the pseudo-residual of an observation $y_t$ of a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.

We also plot histograms of $Y$ and $Z^*$ corresponding to each dive type or subdive state in Figure \ref{fig:empirical_dist}. Observations are weighted by the decoded probability that they correspond to a particular hidden state, and this histogram is plotted over the emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}). Our results again mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed. Another exception is $Z^{*(1)}$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}. See Section S-2.? for histograms for every observation and hidden state.

@@

The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale. However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives. This model therefore does not give any information regarding the dive-level Markov chain or the relationship between the dive and subdive levels. For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives compared to shorter one.

The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3. In addition, the HHMM-DFT produces estimates of $\sigma_1^{*(1)}$, $\sigma_1^{*(2)}$, and $\sigma_1^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those of the CarHHMM-DFT across all axes of acceleration. The estimated uncertainties of the estimates of $\mu_1^{*(1)}$, $\mu_1^{*(2)}$, and $\mu_1^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration. These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation. Further, the empirical auto-correlation of $Z^{*(1)}$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers. Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.

Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it fails produce biologically interpretable results when decoding the subdive states. This fact alone essentially disqualifies the CarHHMM as a viable model for this data. The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution. See section S-2 for a more complete set of results for each of the candidate models.

### 2020-07-31 Report/Sections/simulation.tex
Stylistic changes to report

To compare the performance of the candidate models when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT model. The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}. The ground-truth parameters differ slightly from those in the case study for primarily pedagogical reasons. For example, we set $\Gamma$ such that the simulated data contains approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states is set to $N^*=2$ to simplify the analysis of the subdive states, and $Z^{*(2)}$ is changed from a vector to a scalar for additional simplicity.

We fit all four candidate models to the data generated from the CarHHMM-DFT to analyze the impact of model mis-specification and compare them using several different metrics. These metrics include the decoding accuracy of the dive types and subdive states, bias in parameter estimates, standard error of parameter estimates, and fitting times. We also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the observed Fisher information under that model. This allows us to judge how accurate each model is when estimating its own uncertainty.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive is one of $N=2$ dive types and the coarse-scale probability transition matrix is set as
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are set as
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After the dive durations are calculated for all 100 dives for each data set using $\Gamma$ and $\Theta$, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model. The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive are discarded. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
The observations $Z^*$ are generated using the following parameters:
%
\begin{gather*}
    \mu_1^{*(1)} = 0.0 s, \enspace \sigma_1^{*(1)} = 0.05s, \enspace \phi_1^{*(1)} = 0.99, \\
    %
    \mu_1^{*(2)} = 0.0 s, \enspace \sigma_1^{*(2)} = 0.1s, \enspace \phi_1^{*(2)} = 0.95, \\
    %
    \mu_2^{*(1)} = 10.10, \quad \sigma_2^{*(1)} = 3.18, \\
    %
    \mu_2^{*(2)} = 305.94, \quad \sigma_2^{*(2)} = 17.49.
\end{gather*}
%
Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe a method to reconstruct one possible realization of $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.

The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher in subdive state 2. This corresponds to much more vigorous and variable periodic behavior in the acceleration data.

@@

As expected, the full CarHHMM-DFT is the best-performing model. Given the true subdive type of a particular segment $X^*_{t,t^*}$, the CarHHMM-DFT predicts the correct subdive type with an average confidence of $1 - 10^{-12}$ across all simulations. For the dive types, this average confidence was above 0.9 (see Table \ref{table:accuracy}). Estimates of hidden states are made using the forward-backward algorithm. All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased. Even still, these biases are the least severe for the CarHHMM-DFT among the four models. The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error. See Section S-1.2, tables 2-5 for more information.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. The HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$. For both subdive types, $\hat \sigma_1^*$ greatly overestimates the true standard deviation. The empirical standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ are both much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3). This implies that parameter estimates can be too confident when auto-correlation is ignored.

The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types. The CarHHMM does not have access to the Fourier sums, which are the most informative observations in terms of determining subdive type. It is much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1). The CarHHMM is also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy. $\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates under the CarHHMM (see Section S-1.1.3).

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, it lacks any hierarchical structure, so it is unable to estimate dive types. The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of \textit{all} dives. These parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds. See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.

>>>>

To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT. The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}. The ground-truth parameters differ slightly from those learned in the case study primarily for simplicity of presentation. For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states is set to $N^*=2$ and $Z^{*(2)}$ is set as a scalar for additional simplicity. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times. We also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model. This allows us to judge how accurate each model is when estimating its own uncertainty.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model. The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive are discarded. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
The observations $Z^*$ are generated using the following parameters:
%
\begin{gather*}
    \mu_1^{*(1)} = 0.0 s, \enspace \sigma_1^{*(1)} = 0.05s, \enspace \phi_1^{*(1)} = 0.99, \\
    %
    \mu_1^{*(2)} = 0.0 s, \enspace \sigma_1^{*(2)} = 0.1s, \enspace \phi_1^{*(2)} = 0.95, \\
    %
    \mu_2^{*(1)} = 10.10, \quad \sigma_2^{*(1)} = 3.18, \\
    %
    \mu_2^{*(2)} = 305.94, \quad \sigma_2^{*(2)} = 17.49.
\end{gather*}
%
Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.

The two dive types differ in that dives of type 1 are much shorter on average than dives of type 2. The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher in subdive state 2. This corresponds to much more vigorous and variable periodic behavior in the acceleration data for subdive state 2.

@@

As expected, the full CarHHMM-DFT is the best-performing model. Using the forward-backward algorithm, the CarHHMM-DFT decodes the correct subdive type given the fine-scale data with an average confidence of $1 - 10^{-12}$ across all simulations. For dive type the average confidence was above 0.9 (see Table \ref{table:accuracy}). All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased. Even still, these biases are the least severe for the CarHHMM-DFT among the four models. The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error. See Section S-1.2, tables 2-5 for more information.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. However, the HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$. For both subdive types, $\hat \sigma_1^*$ greatly overestimates the true standard deviation. The estimated standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Section S-1.2, Table 3). This implies that variance can be overestimated and parameter estimates can be too confident when auto-correlation is ignored.

The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types. This result is consistent with expectations because the CarHHMM does not have access to the ``wiggles" of the fine-scale process, which are the most informative observations when determining subdive type. The CarHHMM is also much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1). The CarHHMM is the worst of the four at estimating parameters, and this is in part due to its relatively poor decoding accuracy. $\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates (see Section S-1.1.3).

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, it lacks any hierarchical structure, so it is unable to estimate dive types. The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of \textit{all} dives. The parameter estimates for this distribution ($\hat \mu$ and $\hat \sigma$) are highly correlated and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds. See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.
